Depression and suicidality as costly signals of need

### 2021-04-12 Paper.rmd
Discussion changes - reasonably large

Of the types of negative life events associated with depression, those which involve conflict tend to be more strongly associated with depression [@Hammen2005; @Mazure1998]. Marital problems, bullying, and the existence of abusive relationships are all commonly cited risk factors for depression [@KendlerEtAl1995; @KendlerEtAl1999; @klomek2007; @klomek2019], with  assault (both sexual and non-sexual), in particular, greatly increasing one's risk of depression [@KendlerEtAl1995; @KendlerEtAl1999].  Other notable risk factors like loss of a loved one, financial problems, job loss, or severe or prolonged illness, might seem less related to conflict. However, the ability for others to provide meaningful support to limit the harm caused in times of need [@SugiyamaSugiyama2003] combined with the ubiquity of conflicts of interests mean that many types of problems will eventually involve social conflict even if they did not start that way [@Hagen2003)]. In line with this reasoning, conflict, in general, was found to be associated with depression among the Tsimane, particularly when it involved non-kin [@StieglitzEtAl2015a].

@@

Due to the ease in which signals can be faked [@Zahavi1975; @mokkonen2016] and the costs of accepting a deceptive signal as reliable [@dawkins1978; @johnstone1993; @Hasson1994], demonstrating that one’s need is genuine has likely been a recurrent adaptive problem since humans evolved the capacity to provide instrumental support to one another. One possible solution to this is signaling in ways that are unprofitable to fake due to the costs involved. [@Zahavi1975]. This is possible whenever the signal in question is comparatively cheaper to produce for those who have the underlying quality relative to those who lack it [@grafen1990]. In these instances, individuals with the quality maximize their fitness at greater levels of investment in the signal compared to those who lack the quality or possess it to a lesser degree [@zahavi1977]. The end result of this is expected to be a signal that either tends to not be worth faking or is impossible to fake to a large degree due to selection against fakers constraining the physiology behind the signal [@zahavi1977; @gangestad2005].

@@

The costs of depression and suicidal behavior may be particularly important in times of conflict or when the level of support needed to address one’s situation is substantial, both of which should be expected to increase the payoff of manipulating others through fake signals. When either condition applies, less costly signals like verbal requests, facial sadness, and crying are likely to be less reliable as their low costs can easily be made up for with increased social support, thereby favoring the use of costlier signals, so long as the costs of signaling are outweighed by the benefits [@SymeHagen2021].

Beyond signaling need to those invested in the well-being of the signaler, depression and suicidal behavior may also help an individual bargain for increased support [@Hagen2003]. This is possible due to the reduced investment in others that frequently accompanies depression or the threat to future cooperation that comes from suicide. In these instances, depression and suicide attempts may function analogously to a labor strike, sending the message that one’s need is genuine and that if the target of the signal wanted to continue receiving benefits from the depressed, they would need to increase support towards the signaler. Due to this bargaining component, aversive responses from the targets of the signal to the signaler are expected throughout the bargaining process, with potential helpers likely signaling the costliness of providing support. Importantly, such responses do not prevent the delivery of meaningful benefits, they are simply expected to reduce the risk of helpers providing more support than would benefit them.

@@

The sole randomized treatment in this study was the signal participants saw, with these signals being one of (1) the control condition: a verbal request without additional signaling, (2) crying, (3) mild-depression, (4) depression, and (5) a suicide attempt, with the conditions being ordered from low to high cost.

@@

To reduce the risk of low quality responses, we used a variety of attention checks, which we clearly labeled in an attempt to avoid demand characteristics. The first was shown immediately after the consent form and provided participants with a random word and asked them to enter the vowels in the order in which they are found in the word. Capitalization did not matter and no word contained repeat vowels. Failure to enter the correct vowels in order resulted in the participants being filtered out of the study. The second followed the vignette and involved asking three questions about the story that should be easy to answer for anyone paying attention.

@@

We also asked which emotions participants felt the signaler was experiencing using a multiple choice question in which they could select multiple options. This included emotions directly related to the signals (e.g., sad, depressed, and suicidal), states which suggest genuine need (traumatized and violated), and states which suggest deception (e.g., deviousness or jealousy). The complete list can be found in the SI.

@@

We asked questions identical to those asked at T1 as the main post-treatment variables of interest (T2). For the belief, action, and divide variables, the position of the slider starting where participants placed it at T1. For the T2 emotion variables this involved no changes from the T1 version.

@@

The final part of the survey was a brief demographic questionnaire which examined: (1) the age, (2) sex, (3) number of siblings, (4) number of sons, (5) number of daughters, (6) current relationship status, (7) highest level of education, and (8) the annual household income of each participant.

@@

While the effect of *Suicide attempt* on belief at T2 was similar to the *Depression* conditions, it resulted in greater likelihood of support. This may be due to our decision to measure belief in need as the degree participants believed the claims of the signaler (i.e., the assaults or incompatibility with a potential husband in the context of arranged marriage) rather than a more inclusive measure. For this reason, signal cost may have provided additional information about the costliness of the signaler's adversity in addition to information relevant to whether the event claimed actually happened.

@@

## Broader discussion

Our results on the connection between belief in need and likelihood of help are consistent with costly signaling and bargaining approaches to depression and suicide. In addition to the effect of signal cost on support being largely mediated by the signal's effect on belief in need, belief in need and likelihood of support were highly correlated but distinct at T1. This situation should arise in any instance when a target of a signal of need would benefit from helping if the need was genuine and known to reasonable precision. Importantly, such problems are common among modern humans and have likely been so since individuals could increase their inclusive fitness through supporting kin or important cooperative partners in times of genuine need [@SugiyamaSugiyama2003; @tooby1996].

Although this study was not designed to examine whether the costs of signaling would be made up for by the increased support stemming from depression and suicide attempts directly, the results are consistent with this hypothesis. In general, the costs required to honestly convey one's qualities should be expected to increase with the potential payoffs of deception [@Rosenthal1993], with the vignettes in this study emphasizing the existence of private information and sizable conflicts of interest to increase the perceived risk of being taken advantage of. In similar real-world scenarios, this difficulty of signaling honest need would coincide with the substantial fitness-relevant benefits of preventing future assaults or being able to marry the partner of one's choice. These benefits coupled with the fact (too strong?) that depression and suicide attempts become relatively less costly with increased need [@Hagen2003], suggest that the increased support seen in this study may be worth the costs required to signal need more often than is often appreciated when counting the entire costs of depression as the signal rather than only the upregulation of symptoms for a communicative purpose.

@@

The need to explain suicide's connection to depression is often overlooked by explanations for major depression, particularly those which propose non-signaling functions. One explanation for their connection, is that both suicidality and depression may function to limit negative fitness outcomes in times of substantial adversity [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], with costlier signals being required to elicit support when the payoffs to faking need are high enough so that less costly signals are not reliable indicators of one's need [@Rosenthal1993]. For this hypothesis to be correct, instrumental support would need to be provided frequently enough with large enough impacts so that the costs of signaling are outweighed by the benefits on average. In support of this, we found exposure to depression and suicide attempts resulted in greater belief in need and increased likelihood of support in fictional scenarios. This coupled with findings that depression can simultaneously elicit both positive and negative responses [@HokansonEtAl1986; @stephens1987], suggests that more attention should be paid to the potential for positive responses depressed individuals, especially from those the depressed are in conflict with and those with the capability to provide meaningful benefits.

@@

# To do:
- add vignettes to SI (Mild depression = crying plus behavior change)
- mention somewhere referring to them all as signals even though we know they provide info as cues too

>>>>

Of the types of negative life events associated with depression, those which involve conflict tend to be more strongly associated with depression [@Hammen2005; @Mazure1998]. Marital problems, bullying, and the existence of abusive relationships are all commonly cited risk factors for depression [@KendlerEtAl1995; @KendlerEtAl1999; @klomek2007; @klomek2019], with  assault (both sexual and non-sexual), in particular, greatly increasing one's risk of depression [@KendlerEtAl1995; @KendlerEtAl1999].  Other notable risk factors like loss of a loved one, financial problems, job loss, or severe or prolonged illness, might seem less related to conflict. However, the ability for others to provide meaningful support to limit the harm caused in times of need [@SugiyamaSugiyama2003] combined with the ubiquity of conflicts of interests mean that many types of problems will eventually involve social conflict even if they did not start that way [@Hagen2003]. In line with this reasoning, conflict, as a whole, was found to be associated with depression among the Tsimane, particularly when it involved non-kin [@StieglitzEtAl2015a].

@@

Due to the ease in which signals can be faked [@Zahavi1975; @mokkonen2016] and the costs of accepting a deceptive signal as reliable [@dawkins1978; @johnstone1993; @Hasson1994], demonstrating that one’s need is genuine has likely been a recurrent adaptive problem since humans evolved the capacity to provide instrumental support to one another. One possible solution to this is signaling in ways that are unprofitable to fake due to the costs involved. [@Zahavi1975]. This is possible whenever the signal in question is comparatively cheaper to produce for those who have the relevant underlying quality relative to those who lack it [@grafen1990]. In these instances, individuals with the quality maximize their fitness at greater levels of investment in the signal compared to those who lack the quality or possess it to a lesser degree [@zahavi1977]. The end result of this is expected to be a signal that either tends to not be worth faking or is impossible to fake to a large degree due to selection against fakers constraining the physiology behind the signal [@zahavi1977; @gangestad2005].

@@

The costs of depression and suicidal behavior may be particularly important in times of conflict or when the level of support needed to address one’s situation is substantial, both of which should be expected to increase the payoff of manipulating others through fake signals. When either condition applies, less costly signals like verbal requests, facial sadness, and crying are likely to be less reliable as their low costs can easily be made up for with increased social support, thereby favoring the use of costlier signals so long as the costs of signaling are outweighed by the benefits [@SymeHagen2021].

Beyond signaling need to those invested in the well-being of the signaler, depression and suicidal behavior may also help an individual bargain for increased support [@Hagen2003]. This is possible due to the reduced investment in others that frequently accompanies depression or the threat to future cooperation that comes from suicide. In these instances, depression and suicide attempts may function analogously to a labor strike, sending the message that one’s need is genuine and that if the target of the signal wanted to continue receiving benefits from the depressed, they would need to increase support towards the signaler. Due to this bargaining component, aversive responses from the targets of the signal to the signaler are expected throughout the bargaining process, with potential helpers likely signaling the costliness of providing support. Importantly, such responses do not prevent the delivery of meaningful benefits, they are simply expected to reduce the risk of helpers providing too much support for too little need.

@@

The sole randomized treatment in this study was the signal participants saw, with these signals being one of (1) the control condition: a verbal request without additional signaling, (2) crying, (3) mild-depression, (4) depression, and (5) a suicide attempt, with the order of the conditions here moving from low to high cost.

@@

To reduce the risk of low quality responses, we used a variety of attention checks, which we clearly labeled in an attempt to avoid demand characteristics on the outcomes of interest. The first was shown immediately after the consent form and provided participants with a random word and asked them to enter the vowels in the order in which they are found in the word. Capitalization did not matter and no word contained repeat vowels. Failure to enter the correct vowels in order resulted in the participants being filtered out of the study. The second followed the vignette and involved asking three questions about the story that should be easy to answer for anyone paying attention.

@@

We also asked which emotions participants felt the signaler was experiencing using a multiple choice question in which they could select as many options as they would like. This included emotions directly related to the signals (e.g., sad, depressed, and suicidal), states which suggest genuine need (traumatized and violated), and states which suggest deception (e.g., deviousness or jealousy). The complete list can be found in the SI.

@@

We asked questions identical to those asked at T1 as the main post-treatment variables of interest (T2). For the *Belief*, *Action*, and *Divide* variables, the position of the slider starting where participants placed it at T1. For the T2 emotion variables this involved no changes from the T1 version.

@@

The final part of the survey was a brief demographic questionnaire which examined: (1) the age, (2) sex, (3) number of siblings, (4) number of sons, (5) number of daughters, (6) current relationship status, (7) highest level of education, and (8) the annual household income of each participant. The full survey can found in the SI.

@@

While the effect of *Suicide attempt* on belief at T2 was similar to the *Depression* conditions, it resulted in greater likelihood of support. This may be due to our decision to measure belief in need as the degree participants believed the claims of the signaler (e.g., if the assaults actually happened) rather than a more inclusive measure. For this reason, signal cost may have provided additional information about the costliness of the signaler's adversity in addition to information relevant to whether the event claimed actually happened.

@@

## Relevance to costly signaling

Overall, our results on the connection between belief in need and likelihood of help are consistent with costly signaling and bargaining approaches to depression and suicide. At T1, *Belief* and *Action* were highly correlated yet distinct; while at T2, the effect of signal on both *Belief* and *Action* was to increase the magnitude of pro-signaler responses, with the effect of signal on *T2Action* being largely mediated by signal's effect on belief in need. This relationship between belief in need and likelihood of support and the importance of signal cost should be expected in scenarios when a target of a signal of need would benefit from helping if the signaler's need was genuine but face substantial net costs if it was not. Importantly, such scenarios appear to be common among modern humans and have likely been since individuals could increase their inclusive fitness through supporting kin or important cooperative partners who could return to providing benefits to supporters once their situation returned to normal [@SugiyamaSugiyama2003; @tooby1996].

Although this study was not designed to examine whether the costs of signaling would be made up for by the increased support stemming from depression and suicide attempts directly, the results are consistent with this hypothesis. In general, the costs required to honestly convey one's qualities should be expected to increase with the potential payoffs of deception and the number of defectors within the broader population [@Rosenthal1993], with the vignettes in this study emphasizing the existence of private information and sizable conflicts of interest to increase the perceived risk of being taken advantage of. In similar real-world scenarios, this difficulty of signaling honest need would coincide with the substantial fitness-relevant benefits of preventing future assaults or being able to marry the partner of one's choice. These benefits coupled with the fact that depression and suicide attempts become relatively less costly to one's fitness prospects with increased need [@Hagen2003], suggest that the increased support seen in this study may be worth the costs required to signal need more often than is often appreciated when counting the entire costs of depression as the signal rather than only the upregulation of symptoms for a communicative purpose.

@@

The need to explain suicide's connection to depression is often overlooked by explanations for major depression, particularly those which propose non-signaling functions. One explanation for their connection, is that both suicidality and depression may function to limit negative fitness outcomes in times of substantial adversity [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], with costlier signals being required to elicit support when the payoffs to faking need are high enough so that less costly signals are not reliable indicators of one's need [@Rosenthal1993]. For this hypothesis to be correct, instrumental support would need to be provided frequently enough with large enough impacts so that the costs of signaling are outweighed by the benefits on average. In support of this, we found exposure to depression and suicide attempts resulted in greater belief in need and increased likelihood of support in fictional scenarios. This coupled with findings that depression can simultaneously elicit both positive and negative responses [@HokansonEtAl1986; @stephens1987], suggests that more attention should be paid to the potential for positive responses towards depressed individuals, especially from individuals who normally receive substantial benefits from the depressed individual.

@@

# To do:
- add vignettes to SI (Mild depression = crying plus behavior change)
- victim or signaler?
